---
title: "STAT115/215 BIO/BST282 HW2"
Subtitle: "Differential Expression Analysis & Sample Classification"
author: "huzongyao"
date: "Due Date: Sun 10/27/2024 midnight "
header-includes:
  - \usepackage{ctex}
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    toc: TRUE
    toc_float:
      collapsed: TRUE
      smooth_scroll: TRUE
    number_sections: TRUE
---

> 项目GitHub地址：<https://github.com/dongguaguaguagua/Homework_2021> 持续更新中。

## Part I: Differential expression

In this HW, we will evaluate the differentially expressed genes and pathways between breast cancer and normal breast tissues. Our collaborator generated RNA-seq on ten pairs of breast tumors and matched adjacent normal tissues, located at /n/stat115/2021/HW2/raw_data1. The experiments were run in two batches, each batch with 5 pairs of samples, and the batch information is provided in batch.csv. We have run Salmon for gene quantification which is provided in Cannon at /n/stat115/2021/HW2/raw_data1/Salmon_results. Remember to convert Ensembl ID to gene symbol, mapping between Ensembl id and gene symbol using "org.Hs.eg.db" package in R.

> 在这个作业中，我们将评估乳腺癌与正常乳腺组织之间的差异表达基因和通路。我们的合作者对十对乳腺肿瘤及其匹配的邻近正常组织进行了RNA测序，数据位于/n/stat115/2021/HW2/raw_data1。实验分为两个批次进行，每个批次包含5对样本，批次信息在batch.csv中提供。我们已使用Salmon进行基因定量，结果存储在/n/stat115/2021/HW2/raw_data1/Salmon_results的Cannon中。请记得将Ensembl ID转换为基因符号，使用R中的“org.Hs.eg.db”包进行Ensembl ID与基因符号之间的映射。

### Problem I.1

Please install the following R/Bioconductor packages. Then try "library(package)" to make sure the package works.

Note: sva package with Combat function is used for batch effect removal;

DESeq2 package is used for differential gene expression analysis;

tximport package is used for importing transcript-level abundance into gene-level matrices for downstream analysis

ggplot2 package is used for general plotting in R;

pheatmap package is used for heatmap plotting;

dplyr package is used for data frame manipulations in R;

fgsea package is used for gene-set enrichment analysis.

```{r libraries, message = FALSE}
library(ggplot2)
library(sva)
library(DESeq2)
library(tximport)
library(dplyr)
library(fgsea)
library(pheatmap)
library(ComplexHeatmap)
library(clusterProfiler)
library(factoextra)
library(FactoMineR)
library(ape)
library(ggtree)
library(Cairo)
library(org.Hs.eg.db)
```


### Problem I.2

For RNA-seq analysis, visualization using principle component analysis (PCA) or hierarchical clustering is an essential step of exploratory analysis to identify potental batch effect. Please import transcript-level TPM values from Salmon output and convert to gene-level TPM values. Perform PCA of the samples using log2 transformed TPM values. Indicate different tissues (tumor or normal, using shape to show) and batch (1 or 2, using color to show) of each sample. Next try to use hierarchical clustering for these samples.

Do the results suggest the presence of a batch effect?

For this question, you will load Salmon output at /n/stat115/2021/HW2/raw_data1/Salmon_results. You also need to read in batch information provided in /n/stat115/2021/HW2/raw_data1/batch.csv. Remember to convert Ensembl ID to gene symbol, mapping between Ensembl id and gene symbol using "org.Hs.eg.db" package in R.

> 对于RNA-seq分析，使用主成分分析（PCA）或层次聚类进行可视化是探索性分析中的一个重要步骤，以识别潜在的批次效应。请从Salmon输出中导入转录本级TPM值，并转换为基因级TPM值。使用log2转换的TPM值对样本进行PCA。使用形状区分不同组织（肿瘤或正常），使用颜色区分批次（1或2）来表示每个样本。接下来，尝试对这些样本进行层次聚类。
>
> 结果是否暗示存在批次效应？
>
> 对于这个问题，您需要加载位于/n/stat115/2021/HW2/raw_data1/Salmon_results的Salmon输出。同时，您还需要读取位于/n/stat115/2021/HW2/raw_data1/batch.csv中的批次信息。请记得将Ensembl ID转换为基因符号，使用R中的“org.Hs.eg.db”包进行Ensembl ID与基因符号之间的映射。

```{r}
# 加载必要的库
library(org.Hs.eg.db)
library(tximport)
library(rtracklayer)
library(ggplot2)
library(sva)

### 1. 读取样本信息

# Directory where your Salmon output files are stored
salmon_dir <- "/home/huzongyao1/HW/HW2/raw_data1/HW2_sf"
# 获取所有样本的文件路径
sf_files <- list.files(salmon_dir, pattern = "*.sf", full.names = TRUE)
names(sf_files) <- c("N1", "N2", "N3", "N4", "N5", "T1", "T2", "T3", "T4", "T5")
```

```{r}
### 2. 建立转录本ID到基因ID的映射 tx2gene

# tx_names 是包含所有 ENSEMBLTRANS ID 的向量
tx_names <- tximport(
    sf_files,
    type = "salmon",
    txOut = TRUE,
)
tx_names <- rownames(tx_names$abundance)
# 去除 TxVersion
tx_names <- sub("\\..*$", "", tx_names)

# 建立转录本ID到基因ID的映射
tx2gene <- bitr(
    tx_names,
    fromType = "ENSEMBLTRANS",
    toType = "SYMBOL",
    OrgDb = org.Hs.eg.db
)
```

```{r}
### 3. 导入Salmon输出

txi <- tximport(
    sf_files,
    type = "salmon",
    tx2gene = tx2gene,
    ignoreTxVersion = TRUE,
    countsFromAbundance = "scaledTPM"
)

# Access gene-level TPM and log-transform
tpm_matrix <- txi$abundance
log2_tpm_matrix <- log2(tpm_matrix + 1)
gene_variances <- apply(log2_tpm_matrix, 1, var)
log2_tpm_filtered <- log2_tpm_matrix[gene_variances != 0, ]
```

```{r}
### 4. 主成分分析：处理批次效应前

# 读取批次信息并进行PCA
batch_info <- read.csv("/home/huzongyao1/HW/HW2/raw_data1/batch.csv")
pca_result <- prcomp(t(log2_tpm_filtered), scale. = TRUE)
pca_data <- as.data.frame(pca_result$x)
pca_data$X <- rownames(pca_data)
pca_data <- merge(pca_data, batch_info, by.x = "X", by.y = "X")

print(summary(pca_result))
```

```{r}
### 5. 绘制PCA图

ggplot(pca_data, aes(x = PC1, y = PC2, shape = tissue, color = as.factor(batch))) +
    geom_point(size = 3, alpha = 0.7) +
    labs(title = "PCA of Gene Expression Data", x = "PC1", y = "PC2") +
    theme_minimal()
```

> 从上图中可以看出，肿瘤和正常组织在前两个主成分中有一定的分离，但是批次效应也很明显（明显两个batch被分开了）因此，我们需要对数据进行批次效应的调整。

```{r}
### 6. 层次聚类

# 计算样本之间的距离矩阵
dist_matrix <- dist(t(log2_tpm_matrix))
# 进行层次聚类
hc <- hclust(dist_matrix)
# 绘制层次聚类树
plot(hc, labels = batch_info$X)
```

### Problem I.3

Run COMBAT on the samples to remove the batch effect. Visualize the results using a similar PCA and hierarchical clustering as Problem 2. Provide evidence that the batch effects are successfully adjusted.

> 运行COMBAT算法对样本进行批次效应的去除。使用与问题2类似的PCA和层次聚类来可视化结果。提供证据证明批次效应已成功调整。

```{r}
### 1. 选择那些通过方差过滤的基因

# 设定一个方差阈值（可以调整），只保留高变异的基因
variance_threshold <- quantile(gene_variances, 0.75) # 选择方差在前25%的基因
filtered_genes <- rownames(log2_tpm_matrix)[gene_variances > variance_threshold]
# 提取这些基因的TPM值矩阵
filtered_tpm_matrix <- log2_tpm_matrix[filtered_genes, ]
```

```{r}
### 2. 使用COMBAT去除批次效应

# 从批次信息中提取批次
batch <- as.factor(batch_info$batch)
# 运行Combat算法，去除批次效应
combat_corrected <- ComBat(
    dat = filtered_tpm_matrix,
    batch = batch,
    par.prior = TRUE,
    prior.plots = FALSE
)
```

```{r}
### 3. 批次效应去除后的PCA

# 对经过Combat调整后的数据进行PCA
combat_pca_result <- prcomp(t(combat_corrected), scale. = TRUE)

# 提取PCA结果
combat_pca_data <- as.data.frame(combat_pca_result$x)
combat_pca_data$X <- rownames(combat_pca_data)
combat_pca_data <- merge(combat_pca_data, batch_info, by.x = "X", by.y = "X")

summary(combat_pca_result)
```

```{r}
### 4. 可视化PCA结果

ggplot(combat_pca_data, aes(x = PC1, y = PC2, shape = tissue, color = as.factor(batch))) +
    geom_point(size = 3) +
    labs(title = "PCA After Batch Effect Removal", x = "PC1", y = "PC2") +
    theme_minimal()
```

> PCA结果为PC1占比44.93%，PC2占比13.12%，这两个主成分共解释了58.05%的方差。
>
> 从上图中可以看出，批次效应被成功去除，肿瘤和正常组织在前两个主成分中有一定的分离，但是不同批次之间的分隔已经消失。

```{r}
### 5. 层次聚类分析

# 计算样本之间的距离矩阵
combat_dist_matrix <- dist(t(combat_corrected))
# 进行层次聚类
combat_hc <- hclust(combat_dist_matrix)
# 绘制层次聚类树
plot(combat_hc, labels = batch_info$X, main = "Hierarchical Clustering After Batch Effect Removal")
```

### Problem I.4

Run DESeq2 based on paired samples adjusting for the batch effect to identify differentially-expressed genes between tumor and normal tissues. How many genes are expressed higher in tumors than normal. Let's use 1) FDR < 0.01 and 2) Log2 fold change > 1 as the cutoff.

Note: please use the raw_count columns of the Salmon result and convert these to integer values for DESeq2.

Identify the top 5 most (by Log2FC) over expressed genes (FDR < 0.01) in tumor and normal, respectively.

> 基于成对样本运行DESeq2，调整批次效应，以识别肿瘤与正常组织之间的差异表达基因。问有多少基因在肿瘤中的表达水平高于正常组织。我们使用以下标准作为筛选条件：
>
> 1. FDR < 0.01
> 2. Log2倍数变化 > 1。
>
> 注意：请使用Salmon结果中的raw_count列，并将其转换为整数值以用于DESeq2分析。
>
> 分别识别在肿瘤和正常组织中表达上调最高的前5个基因（按Log2倍数变化排序，FDR < 0.01）。

```{r}
library(DESeq2)
library(sva)
library(ggplot2)

#### DESeq2 from salmon

# 使用tximport加载Salmon结果
txi_raw <- tximport(
    sf_files,
    type = "salmon",
    tx2gene = tx2gene,
    ignoreTxVersion = TRUE,
    countsFromAbundance = "no" # 使用原始计数(raw_count)，而不是归一化后的TPM值
)

# 提取基因级的raw_count（即NumReads）并确保为整数格式
raw_counts <- round(txi_raw$counts)
print(head(raw_counts, 10))
```

```{r}
### extract differentially expressed genes

# 定义样本分组信息（正常 vs 肿瘤）
col_data <- batch_info
# 将tissue和batch转换为因子
col_data$tissue <- factor(col_data$tissue)
col_data$batch <- factor(col_data$batch)

# 创建DESeq2对象，调整批次效应
dds <- DESeqDataSetFromMatrix(
    countData = raw_counts,
    colData = col_data,
    # Deseq2 中加入batch因子以去除批次效应
    design = ~ batch + tissue
)

# 保证每个基因至少有一个样本有足够的读取数
dds <- dds[rowSums(counts(dds)) > 1, ]
```

```{r}
# 运行DESeq2并进行差异表达分析
dds <- DESeq(dds)

# 提取结果：比较肿瘤 vs 正常，筛选显著性差异表达基因
dds_res <- results(dds, contrast = c("tissue", "Tumor", "Normal"))
# 筛选符合FDR < 0.01和Log2倍数变化 > 1的基因
res_filtered <- dds_res[which(dds_res$padj < 0.01 & abs(dds_res$log2FoldChange) > 1), ]
```

```{r}
# 检查肿瘤中表达水平高于正常组织的基因
upregulated_genes <- res_filtered[which(res_filtered$log2FoldChange > 1), ]
# 提取在肿瘤中表达最高的前5个基因（按Log2倍数变化排序）
top5_upregulated <- head(upregulated_genes[order(-upregulated_genes$log2FoldChange), ], 5)

# 检查肿瘤中表达水平低于正常组织的基因
downregulated_genes <- res_filtered[which(res_filtered$log2FoldChange < -1), ]
# 提取在正常组织中表达最高的前5个基因（按Log2倍数变化排序）
top5_downregulated <- head(downregulated_genes[order(downregulated_genes$log2FoldChange), ], 5)

# 输出在肿瘤中的表达水平高于正常组织的基因数量
cat("Number of genes upregulated in tumor:", nrow(upregulated_genes), "\n")

# 输出前5个基因的基因符号和Log2FoldChange
cat("Top 5 upregulated genes in tumor:\n")
print(top5_upregulated[, c("log2FoldChange", "padj")])

cat("\nTop 5 downregulated genes in normal tissue:\n")
print(top5_downregulated[, c("log2FoldChange", "padj")])
```

```{r}
# 提取前5个上调基因

# 提取这些基因的原始计数
top5_combined_counts <- raw_counts[
    rownames(raw_counts) %in% union(
        rownames(top5_upregulated),
        rownames(top5_downregulated)
    ),
]

# 计算肿瘤和正常组织中的平均表达量
avg_counts <- data.frame(
    gene = rownames(top5_combined_counts),
    normal = rowMeans(top5_combined_counts[, col_data$tissue == "Normal"]),
    tumor = rowMeans(top5_combined_counts[, col_data$tissue == "Tumor"])
)

# 将数据转换为长格式
library(tidyr)
avg_counts_long <- avg_counts %>%
    pivot_longer(
        cols = c(normal, tumor),
        names_to = "tissue",
        values_to = "count"
    )

# 绘制前5个基因的柱状图
ggplot(avg_counts_long, aes(x = reorder(gene, -count), y = count, fill = tissue)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(title = "Top 5 Upregulated Genes in Tumor vs Normal", x = "Gene", y = "Average Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Problem I.5

Visualize the differential gene expression values by making a volcano and an MA plot to summarize the differences between tumor and normal. In the volcano plot, draw the statistically significant (FDR < 0.01, Log2FC > 1) tumor up-genes red and down-genes blue.

Note: Be sure to use the lfcShrink function to get more robust estimates of the fold-changes for genes.

> 可视化基因差异表达值，通过火山图和MA图总结肿瘤与正常之间的差异。在火山图中，将具有统计显著性的基因（FDR < 0.01，Log2FC > 1）显示为红色的上调基因和蓝色的下调基因。
>
> 注意：请确保使用 `lfcShrink` 函数，以获得基因倍数变化的更稳健估计。

```{r}
### lfcShrink to get more robust estimates of fold changes for genes

# 使用 apeglm 方法收缩 Log2FC，确保结果更稳健
res_shrunken <- lfcShrink(dds, coef = "tissue_Tumor_vs_Normal", type = "apeglm")

### Volcano plot

# 转换为数据框以便使用 ggplot2
res_shrunken_df <- as.data.frame(res_shrunken)

# 添加颜色标记列：红色表示上调基因，蓝色表示下调基因，灰色表示无显著性差异基因
res_shrunken_df$color <- "Non-significant"
res_shrunken_df$color[res_shrunken_df$padj < 0.01 & res_shrunken_df$log2FoldChange > 1] <- "Up Regulated Genes"
res_shrunken_df$color[res_shrunken_df$padj < 0.01 & res_shrunken_df$log2FoldChange < -1] <- "Down Regulated Genes"

# 绘制火山图
ggplot(res_shrunken_df, aes(x = log2FoldChange, y = -log10(padj), color = color)) +
    geom_point(alpha = 0.8, size = 1.5) +
    scale_color_manual(values = c(
        "Down Regulated Genes" = "blue",
        "Up Regulated Genes" = "red",
        "Non-significant" = "grey"
    )) +
    labs(title = "Volcano Plot", x = "Log2 Fold Change", y = "-Log10 Adjusted P-value") +
    theme_minimal()
```

```{r}
# 绘制 MA 图
plotMA(res_shrunken, ylim = c(-5, 5), main = "MA Plot")
```

### Problem I.6

Try kmeans (try k = 4 or 7) clustering to group differentially expressed genes into different clusters. How many genes are there in each cluster? Draw a heatmap showing gene clusters.

> 尝试使用 k-means 聚类（尝试 k = 4 或 7）对差异表达基因进行分组。每个聚类中有多少个基因？绘制展示基因聚类的热图。

```{r}
# kmeans for differential genes

# 提取差异表达基因的表达矩阵
gene_data <- assay(dds)[rownames(res_filtered), ]
# 标准化基因表达数据（通常使用行标准化）
gene_data_scaled <- t(scale(t(gene_data)))
# 设置聚类数 k
k <- 7
# 进行 k-means 聚类
set.seed(114514) # 设置随机种子以获得可重复结果
kmeans_result <- kmeans(gene_data_scaled, centers = k)
```

```{r}
# summarize gene number in each cluster

# 查看每个聚类中的基因数量
table(kmeans_result$cluster)
```

```{r}
# heatmap showing genes with kmeans clustering

# 使用 pheatmap 包绘制热图
library(pheatmap)
# 使用 k-means 聚类结果对基因重新排序
gene_data_ordered <- gene_data_scaled[order(kmeans_result$cluster), ]
# 创建热图并添加聚类标签
pheatmap(gene_data_ordered,
    cluster_rows = FALSE, cluster_cols = TRUE,
    annotation_row = data.frame(Cluster = factor(kmeans_result$cluster)),
    show_rownames = FALSE, show_colnames = TRUE,
    main = paste("Heatmap of Differentially Expressed Genes (k =", k, ")")
)
```

### Problem I.7: For graduate students only

If you run DESeq2 without removing batch effect, how many differential genes do you get? How do their K-means clustering look? Does batch effect removal gives more interpretable results?

> 如果在不去除批次效应的情况下运行 DESeq2，你得到多少差异基因？它们的 K-means 聚类结果如何？去除批次效应后，结果是否更具可解释性？

> 但是，在尝试不去除批次效应的情况下运行 DESeq2 前，我想使用 `limma::removeBatchEffect` 函数进一步校正批次效应，并比较两种方法的效果。

```{r}
library(limma)
library(stats)

# 使用 limma::removeBatchEffect 进一步校正批次效应
adjusted_gene_data <- removeBatchEffect(gene_data, batch = col_data$batch)

# 进行 PCA 分析
adjusted_pca_res <- prcomp(t(adjusted_gene_data), scale. = TRUE)

# 提取前两个主成分
adjusted_pca_data <- as.data.frame(adjusted_pca_res$x)
adjusted_pca_data$X <- rownames(adjusted_pca_data)
adjusted_pca_data <- merge(adjusted_pca_data, col_data, by.x = "X", by.y = "X")

print(summary(adjusted_pca_res))

# 绘制 PCA 图
ggplot(adjusted_pca_data, aes(x = PC1, y = PC2, shape = tissue, color = as.factor(batch))) +
    geom_point(size = 3, alpha = 0.7) +
    labs(title = "PCA of Gene Expression Data", x = "PC1", y = "PC2") +
    theme_minimal()
```

```{r}
# 对校正后的数据运行 K-means 聚类

# 标准化基因表达数据（通常使用行标准化）
adjusted_gene_data_scaled <- t(scale(t(adjusted_gene_data)))
# 设置聚类数 k
k <- 7
# 进行 k-means 聚类
set.seed(114514) # 设置随机种子以获得可重复结果
adjuested_kmeans_result <- kmeans(adjusted_gene_data_scaled, centers = k)
# 查看每个聚类中的基因数量
table(adjuested_kmeans_result$cluster)
# 使用 k-means 聚类结果对基因重新排序
adjusted_gene_data_ordered <- adjusted_gene_data_scaled[order(adjuested_kmeans_result$cluster), ]
# 创建热图并添加聚类标签
pheatmap(adjusted_gene_data_ordered,
    cluster_rows = FALSE, cluster_cols = TRUE,
    annotation_row = data.frame(Cluster = factor(adjuested_kmeans_result$cluster)),
    show_rownames = FALSE, show_colnames = TRUE,
    main = paste("[limma adjusted] Heatmap of Differentially Expressed Genes (k =", k, ")")
)
```

```{r}
#### DESeq2 from salmon without batch removal

# 创建DESeq2对象
batched_dds <- DESeqDataSetFromMatrix(
    countData = raw_counts,
    colData = col_data,
    design = ~tissue
)

# 保证每个基因至少有一个样本有足够的读取数
batched_dds <- batched_dds[rowSums(counts(batched_dds)) > 1, ]
# 运行DESeq2并进行差异表达分析
batched_dds <- DESeq(batched_dds)

### extract differentially expressed genes

# 提取结果：比较肿瘤 vs 正常，筛选显著性差异表达基因
batched_dds_res <- results(batched_dds, contrast = c("tissue", "Tumor", "Normal"))
# 筛选符合FDR < 0.01和Log2倍数变化 > 1的基因
batched_res_filtered <- batched_dds_res[which(batched_dds_res$padj < 0.01 & abs(batched_dds_res$log2FoldChange) > 1), ]
```

```{r}
# kmeans for differential genes

# 提取差异表达基因的表达矩阵
batched_gene_data <- assay(batched_dds)[rownames(batched_res_filtered), ]
# 标准化基因表达数据（通常使用行标准化）
batched_gene_data_scaled <- t(scale(t(batched_gene_data)))
# 设置聚类数 k
k <- 7
# 进行 k-means 聚类
set.seed(114514) # 设置随机种子以获得可重复结果
batched_kmeans_result <- kmeans(batched_gene_data_scaled, centers = k)

# summarize gene number in each cluster

# 查看每个聚类中的基因数量
table(batched_kmeans_result$cluster)

# heatmap showing genes with kmeans clustering

# 使用 k-means 聚类结果对基因重新排序
batched_gene_data_ordered <- batched_gene_data_scaled[order(batched_kmeans_result$cluster), ]
# 创建热图并添加聚类标签
pheatmap(batched_gene_data_ordered,
    cluster_rows = FALSE, cluster_cols = TRUE,
    annotation_row = data.frame(Cluster = factor(batched_kmeans_result$cluster)),
    show_rownames = FALSE, show_colnames = TRUE,
    main = paste("[batch] Heatmap of Differentially Expressed Genes (k =", k, ")")
)
```

### Problem I.8

From the batch-removed DESeq2 run results, extract the top 200 tumor-upregulated genes (by Log2FC, FDR < 0.01). Run DAVID GO analysis (http://david.abcc.ncifcrf.gov/) to see whether these genes are enriched in specific biological process (BP), pathways, etc.

> 从去除批次效应的 DESeq2 运行结果中，提取前 200 个肿瘤上调基因（按 Log2FC 排序，FDR < 0.01）。运行 DAVID GO 分析 (http://david.abcc.ncifcrf.gov/) 以查看这些基因是否在特定的生物过程 (BP)、通路等方面富集。

```{r, fig.width=10}
## extract top 200 genes from up-regulated genes with batch removal
top200_upregulated <- head(upregulated_genes[order(-upregulated_genes$log2FoldChange), ], 200)
print(top200_upregulated)

## write gene list to file
write.table(
    rownames(top200_upregulated),
    file = "/home/huzongyao1/Homework_2020/HW2/output/top200_upregulated_genes.txt",
    quote = FALSE,
    row.names = FALSE,
    col.names = FALSE
)
```



### Problem I.9: For graduate students only

Run Gene Set Enrichment analysis (http://www.broadinstitute.org/gsea/index.jsp) using the summary statistics from problem 4. Show the top five gene sets or pathways that best capture the differentially expressed genes between tumor than normal. Comment on the biological relevance of the results. Plot GSEA enrichment plot for an interesting pathway.

Mapping between gene sets and pathways is provided in /n/stat115/2021/HW2/raw_data1/c2.cp.kegg.v7.4.symbols.gmt file file.

```{r, fig.width=10}
#
### GSEA
## prepare pre-ranked gene list


## read in gmt file


## GSEA


## extract significant pathways


## plot GSEA table


## plot GSEA enrichment plot for an interesting pathway
```

```{r, fig.width=10}
### GSEA
## prepare pre-ranked gene list


## read in gmt file


## GSEA


## extract significant pathways


## plot GSEA table


## plot GSEA enrichment plot for an interesting pathway
```

## Part II: Sample classification

We provide you z-score normalized expression data of 50 breast tumor samples, 50 normal breast samples (your training and cross-validation data), and 20 samples without diagnosis (your testing data). We want to use the 100 samples with known diagnosis to train machine learning models in order to predict the 20 unknown samples.

You will need the following libraries in R: ggplot2 and ggfortify for plotting, MASS and caret for machine learning, and pROC is for evaluating testing performance. The YouTube video on caret (https://youtu.be/z8PRU46I3NY) and the package documentation (http://topepo.github.io/caret/index.html) might be helpful.

All data for Part II are provided at /n/stat115/2021/HW2/raw_data2.

```{r, warning=FALSE, cache=FALSE, message=FALSE}
# library(ggplot2)
# library(ggfortify)
# library(pROC)
# library(caret)

#### read in data
```

### Problem II.1

Run PCA for dimension reduction on the 100 samples with known labels, and draw these 100 samples in a 2D plot. Do cancer and normal separate from the first two PCs? Would this be sufficient to classify the unknown samples?

z-score normalized data are provided in BRCA_zscore_data.txt. Phenotype data is in BRCA_phenotype.txt.

```{r, warning=FALSE, cache=FALSE, message=FALSE}
## run PCA

## view the first two PCs
```

We have plotted here a PCA for the training dataset containing expression data from 50 tumor breast samples and 50 normal breast samples. It is important that we centered and scaled the data before conducting the PCA analysis because PCA is a function that maximizes the variance of the chosen components. Since different genes, for instance, have diffewrent variances in their expression biologically, the initial scaling ensures that PCA can more accurately capture the components that truly reflect the underlying variance of the data. If we had run the PCA without scaling first, PC1 and PC2 only explain 19.84% and 13.84% of the variance compared to the 53.33% and 7.58% shown above. Thus, scaling prevented some genes with large variances from dominating the first principle component in our results.

With the PCA plot colored above by phenotype, we can clearly see that cancer and normal tissue separate on the first two principle components given that each color largely clusters with itself. More specifically, we see that this separation is largely produced by PC2. As a rough estimate, normal samples tend to be PC2 < 0.0 while tumor samples are largely PC2 > 0.0. Although we do see separation from the first two PCs, this information would not be sufficient to classify unknown samples because PCA is purely a dimensional reduction tool. As it does not rely on any information provided by sample labels, it is a form of unsupervised machine learning and cannot be used to make predictions as required of a classifier. If we wanted to use these PCA results to make predictions of cancer versus normal, we would first have to use a supervised learning approach by fitting a classifier onto the PCA-transformed data containing class labels. Thus, PCA cannot be used for classification purposes.

### Problem II.2

Draw a plot showing the cumulative % variance captured from the top 100 PCs. How many PCs are needed to capture 90% of the variance?

```{r, warning=FALSE, cache=FALSE, message=FALSE}

```

```{r, warning=FALSE, cache=FALSE, message=FALSE}
# How many PCs are needed to capture 90% of the variance?
```

In our graph above, we have included a dashed red line to show 90% of the variance. Based on the cumulative sum data, 25 PCs are needed to capture 90% of the variance.

### Problem II.3

Apply machine learning methods (KNN, logistic regression, Ridge regression, LASSO, ElasticNet, random forest, and support vector machines) on the top 25 PCs of the training data and 5-fold cross validation to classify the samples. caret and MASS already implemented all of the machine learning methods, including cross-validation, so calling each is only one command. In order to get consistent results from different runs, use set.seed().

```{r, warning=FALSE, cache=FALSE, message=FALSE}
# input.reduction <- data.frame(pca.res$x[,1:25])
# input.reduction$phenotype <- as.factor(phenotype)
# rownames(input.reduction) <- rownames(pca.res$x)

#### Run models using 7-fold cross validation

# a) basic algorithms

# b) advanced algorithms
# GLM
```

### Problem II.4

Summarize the performance of each machine learning method, in terms of accuracy and kappa.

```{r, warning=FALSE, cache=FALSE, message=FALSE}
# compare accuracy of models(use accuracy and kappa index to measure the performance of these models)
```

The accuracy and kappa statistics for all the machine learning methods used from KNN to SVM have been summarized in the table and dotplot above.

### Problem II.5: For Graduate students only

Compare the performance difference between logistic regression, Ridge, LASSO, and ElasticNet. In LASSO, how many PCs have non-zero coefficient? What is the lambda for Ridge and LASSO, respectively?
```{r, warning=FALSE, cache=FALSE, message=FALSE}
# compare accuracy of models(use accuracy and kappa index to measure the performance of these models)
```


```{r, warning=FALSE, cache=FALSE, message=FALSE}

```

Accuracy is the percentage of correctly classified instances. Kappa is metric based on accuracy that has been normalized against the random chance of seeing each category based on the dataset. When we specifically compare logistic regression, Ridge, LASSO and ElasticNET on these two metrics, we can see that ElasticNET appears to outperform the other models. This result is more apparent when we compare ElasticNET to Ridge and LASSO, where the mean accuracies are 0.980, 0.950 and 0.947 respectively while the mean Kappa values are 0.96, 0.86 and 0.82. The superior performance of ElasticNET is consistent with the underlying model because all three are trying to minimize the residual sum of squares with a penalty term. This penalty term has two tunable parameters - alpha and lambda. Importantly, we tuned both alpha and lambda in order to generate the most accurate model when we applied ElasticNET. By comparison, alpha must be set to 0 for Ridge regression and alpha must be 1 for LASSO, meaning that only lambda can be tuned. Given that we can optimize two parameters in ElasticNET compared to only one in both Ridge and LASSO, it is to be expected that ElasticNET outperforms both Ridge and LASSO. In logistic regression, we are also minimizing the residual sum of squares, but without applying any penalty term. We see here that logistic regression appears to outperform Ridge and LASSO based on the accuracy and kappa summary statistics. The normal purpose of the penalty term is to introduce some bias in order to reduce the variance and avoid overfitting to the training data. It is important here that we see perfect separation in our data, which prevents model convergence and hence the error message. However, this perfect convergence means that the performance of our model will be particularly strong because our model can easily fit the training data. Thus, I am not surprised to see that the penalty introduced in LASSO and Ridge decrease the accuracy and kappa when compared to the perfect convergence that occured in our logistic regression model.

In LASSO, there are 2 non-zero coefficients. However, one of these coefficients is the intercept while the other is PC2. Thus, there is only 1 PC with a non-zero coefficient in LASSO. In ElasticNet, the lambda for Ridge (defined as alpha = 0) is 2.9 while the lambda for LASSO (defined as alpha = 1) is 0.1.

### Problem II.6

Use the PCA projections in Q1 to obtain the first 25 PCs of the 20 unknown samples. Use one method that performs well in Q4 to make predictions. Caret already used the hyper-parameters learned from cross-validation to train the parameters of each method on the full 100 training data. You just need to call this method to make the predictions.

Expression data for the 20 unknown samples are provided in unknown_samples.txt.

```{r, warning=FALSE, cache=FALSE, message=FALSE}
# Use the PCA projections in Q1 to obtain the first 25 PCs
```

The predictions for the 20 test samples were determined by our trained SVM model after projecting the test data onto the eigenvectors of the PCs determined for the training data.

### Problem II.7: For Graduate students only

Can you find out the top 3 genes that are most important in this prediction method in Q6? Do they have some known cancer relevance?

```{r, warning=FALSE, cache=FALSE, message=FALSE}

```

The top three genes in this prediction method are: EXOSC1, LCMT1 and MEF2D. EXOSC1 encodes the exosome complex component CSL4, a core component of the exosome, which is an intracellular complex responsible for RNA degradation. LCMT1 encodes leucine carboxyl methyltransferase 1, which catalyzes methylation of protein phosphatase-2A. MEF2D encodes myocyte enhancer factor 2D, which is a transcription factor involved in muscle and neuronal differentiation. Based on gene ontology analysis through DAVID to see the functional annotations of all three genes, we get terms consistent with these functions. EXOSC1 returns RNA degradation as a KEGG pathway while LCMT1 returns histidine metabolism and estrogen metabolism, indicating a mainly metabolic function. MEF2D returns skeletal system development and nucleotide binding.

In the context of cancer, EXOSC1 has been reported as a poor prognostic marker in renal cancer and to have low cancer specificity. Similarly, there are no studies definitively linking LCMT1 expression to cancer. For MEF2D, one recent study suggests that a pathway between MEF2D and Wnt signalling are linked to cancer proliferation, but there is no large or convincing body of evidence. The clear lack of a strong biological link between these three genes and cancer is not surprising. Our machine learning models are completely blind to biological function and trained to pick the genes that provide the best accuracy. Hence, it is clear here that the top three genes found to be important in our SVM model do not have biologically established roles in cancer despite being useful for classification purposes.

### Problem II.8

Suppose a pathologist later made diagnosis on the 20 unknown samples (load the diagnosis.txt file). Based on this gold standard, draw an ROC curve of your predictions in Q6. What is the prediction AUC?

```{r, warning=FALSE, cache=FALSE, message=FALSE}
# load the diagnosis file
## ROC
```

The prediction AUC is 0.879, suggesting that our classifier is fairly good classifier. It correctly classifies samples 87.9% of the time, so it is not perfect (AUC = 1), but far better than chance (AUC = 0.5).
